---
title: "sidtoolbox"
output: html_document
author: Marzie Rasekh
---
# Subgroup discovery demonstration

```{r load_packaged, message=FALSE, warning=FALSE, include=FALSE}

# or by indices
library(devtools)
library(roxygen2)

#install(".")
# or load it
library(sidtoolbox)
# update if changes made
document()
library(data.table)
library(stringr)
library(ggplot2)

```

# Simulated data


```{r simulate_data}

# simulate a sample dataset with 3 types of variables, numeric, binary, and count
# covariates are X1, X2, ...
# outcomes are Y_[type], namely Y_numeric, Y_binary, and Y_count
# the contrast arm is TRT
# 'N' samples will be created 
# also places two subgroups, oneis true positive (random_TP) of size 'subgroup_ratio'*'N' where the treatment arm is enhanced by 'subgroup_enhanced_effect'
# and false positive (random_FP) of same size but where both arms are shifted by 'subgroup_shift_effect'
simulated.twoarm <- simulateSubgroupData(
                            N = 2000,
                            case = "prognostic", 
                            arm = "two",
                            overal_treatment_effect = 0.5, 
                            subgroup_enhanced_effect = 1.5, 
                            subgroup_ratio = 0.1,
                            covariates_normal = list(c(0,1), c(0,1), c(0,1)),
                            correlation = c(),
                            covariates_binary = c(0.5, 0.5, 0.5, 0.5), 
                            covariates_uniform = list(c(0,1), c(0,1), c(0,1)), 
                            has_FP = TRUE,
                            SD_NOISE_RATIO = 0.1)
#names(simulated.twoarm)
# summary(simulated.twoarm$data)
simulated.twoarm

```


## Explore the data
This step helps identify covariates to be included in the subgroup discovery.

### Correlation matrix
Correlations who one on one relation in your data. Highly correlated covariates should be removed.

```{r exploreCorrelationMatrix}

# examples
exploreTriangleCorrelationMatrix(subgroup.data = simulated.twoarm, method = "spearman")
exploreCorrelationMatrix(subgroup.data = simulated.twoarm)
exploreDetailedChartCorrelation(subgroup.data = simulated.twoarm)


```

### Association rules
Association rules help us find high level correlations in our data. 
The rules are controled by support (how large the corresponding subgroup is) and confidence (how many of the total subgroup size follow the rule).

```{r exploreAssociationRules, message=FALSE, warning=FALSE}

# res <- exploreAssociationRules(subgroup.data)
listAssociationRules(subgroup.data = simulated.twoarm, onlyCovariates = FALSE)
plotAssociationRules(subgroup.data = simulated.twoarm, onlyCovariates = TRUE)
plotAssociationRules(subgroup.data = simulated.twoarm, method = "graph", measure = "confidence", onlyCovariates = TRUE)
plotAssociationRules(subgroup.data = simulated.twoarm, method = "scatterplot", measure = "support", onlyCovariates = TRUE)
plotAssociationRules(subgroup.data = simulated.twoarm, method = "two-key plot", measure = "support", onlyCovariates = FALSE)
#plotAssociationRulesInteractive(subgroup.data = simulated.twoarm, onlyCovariates = TRUE)


```

### Conditional inference tree
Conditional inference trees are useful to fund the false positive subgroups, where both arms are shifted.


```{r exploreDecisionTree, message=FALSE, warning=FALSE}
# the true positive / false positive subgroup is
# subgroup.data$rules
a <- sapply(exploreConditionalInferenceTree(subgroup.data = simulated.twoarm, stump = FALSE, testtype = "Bonferroni", teststat = "max", significance = 0.05, nmin = 0.05, maxdepth = 2), FUN = plot)

```



### Clustering 


```{r clustering}
# coming soon

```

## Subgroup discovery

```{r subgroup_discovery}
# set the outcomes you want to run subgroup discovery on
outcome_columns <- colnames(simulated.twoarm$data)[simulated.twoarm$outcomes]
# save all rules in allRules
allRules <- simulated.twoarm$rules
allRules

```

### Treatment-specific Subgroup Discovery Tool (TSDT)
<b>
Battioui, C., Shen, L., Ruberg, S., (2014). <i>A Resampling-based Ensemble Tree Method to Identify
Patient Subgroups with Enhanced Treatment Effect.</i> JSM proceedings, 2014 </b>

```{r TSDT, message=FALSE, warning=FALSE}
# the true positive / false positive subgroup is
# simulated.twoarm$rules

param <- list(
  desirable_response = "increasing",
  nsamples = 100, 
  npermutations = 100, 
  maxdepth = 2, 
  min_subgroup_n_control = 0.05, 
  min_subgroup_n_trt = 0.05, 
  n_cpu = 1
)

resTSDT <- lapply(X = outcome_columns, FUN = function(y_idx) {
  res <- runTSDT(subgroup.data = simulated.twoarm, 
                     y_idx = y_idx,
                     desirable_response = param$desirable_response,
                     nsamples = param$nsamples, 
                     npermutations = param$npermutations, 
                     maxdepth = param$maxdepth, 
                     min_subgroup_n_control = param$min_subgroup_n_control, 
                     min_subgroup_n_trt = param$min_subgroup_n_trt, 
                     n_cpu = param$n_cpu)
  res
  allRules <<- rbind(allRules, 
                    parseTSDTResults(TSDT_table = res, outcome = y_idx, param = param))
})

resTSDT

```

### Virtual Twins

Currently only supports binary outcomes and two arm studies.

<b>Foster, Jared C., Jeremy MG Taylor, and Stephen J. Ruberg. <i>Subgroup identification from randomized clinical trial data.<i> Statistics in medicine 30.24 (2011): 2867-2880. 
<a href="https://www.ncbi.nlm.nih.gov/pubmed/21815180">link</a></b>

<a href="https://cran.r-project.org/web/packages/aVirtualTwins/index.html">Link to implemented package</a>


```{R virtual_twins, message=FALSE, warning=FALSE}
# the true positive / false positive subgroup is
# simulated.twoarm$rules

param <- list(
  desirable_response = "increasing", 
  forest.type = "double",
  method = "absolute", 
  tree.type = "class", 
  folds = 10, 
  ntree = 1000,
  maxdepth = 2
)
idx <- sapply(X = outcome_columns, FUN = function(y){
  x <- simulated.twoarm$data[, y, with = FALSE][[1]]
  is.factor(x) && 
    length(levels(x) == 2)
})
binary_outcomes <- outcome_columns[idx]
resVT <- 
  lapply(X = binary_outcomes,
       FUN = function(y_idx) {
  res <- runVirtualTwins(subgroup.data = simulated.twoarm, 
                         y_idx = y_idx,
                         desirable_response = param$desirable_response, 
                         forest.type = param$forest.type,
                         method = param$method, 
                         tree.type = param$tree.type, 
                         folds = param$folds, 
                         ntree = param$ntree,
                         maxdepth = param$maxdepth)
  allRules <<- rbind(allRules,
                    parseVTResults(VT_table = res, outcome = y_idx, param = param))
  res
})

resVT

```

### Optimization approach with Particle Swarm Optimization (PSO)

```{r PSO}
# the true positive / false positive subgroup is
# simulated.twoarm$rules

resPSO <- lapply(X = outcome_columns,
                 FUN = function(y_idx){
  pso <- runPSO(subgroup.data = simulated.twoarm, 
                y_idx = y_idx, 
                fitness_method = "model", 
                desirable_response = "increasing", 
                depth = 2, 
                nmin = 0.1, 
                iterations = 100)
  allRules <<- rbind(allRules, 
                     pso)
  pso
})
resPSO



```


## Analyze and compare rules

```{r compare_rules}
MIN_P_VALUE <- 10e-10
P_THRESHOLD <- -log10(0.05)
summarizeRules <- function(subgroup.data, rules) {
  N.rules <- nrow(rules)
  summaryRules <- data.table()
  
  allOutcomes <- outcome_columns
  
  for (i in c(1:N.rules)) {
    row <- rules[i]
    # adding a fix to remove dummy rules wihch were negative controls
    if (!grepl(x = row$rule, pattern = "dummy")) {
      if (str_length(row$outcome) == 0) {
        Y <- allOutcomes
      } else {
        Y <- row$outcome
      }
      for (y in Y) {
        row$outcome <- y
        summaryRules <- rbind(summaryRules, row)
      }
    }
    
  }
  N.rules <- nrow(summaryRules)
  N <- nrow(simulated.twoarm$data)
  effect_size <- sapply(
    X = c(1:N.rules),
    FUN = function(i) {
      row <- summaryRules[i]
      sbg_idx <-
        parseRule(covariates = simulated.twoarm, rule = row$rule)
      f <- interactionEffectSize(
        subgroup.data = simulated.twoarm,
        y_idx = row$outcome,
        subgroup_idx = sbg_idx
      )
      c(
        N_ratio = sum(sbg_idx) / N,
        effect = f[1],
        pvalue = f[2]
      )
    }
  )
  summaryRules$N_ratio <- effect_size[1, ]
  summaryRules$interaction_coeff <- effect_size[2, ]
  summaryRules$p_value <- effect_size[3, ]
  summaryRules$p_value[summaryRules$p_value < MIN_P_VALUE] <- MIN_P_VALUE
  # check TP rule and get accuracy precision recall
  truth_subgroup <- which(rules$method == "sim_TP")
  if (length(truth_subgroup) > 0) {
    truth_subgroup <- parseRule(covariates = simulated.twoarm,
                                rule = rules[truth_subgroup[1], ]$rule)
    performance <- sapply(
      X = c(1:N.rules),
      FUN = function(i) {
        row <- summaryRules[i, ]
        sbg_idx <-
          parseRule(covariates = simulated.twoarm, rule = row$rule)
        TP <- sum(sbg_idx &  truth_subgroup)
        FP <- sum(sbg_idx & !truth_subgroup)
        TN <- sum(!sbg_idx & !truth_subgroup)
        FN <- sum(!sbg_idx &  truth_subgroup)
        precision <- TP / (TP + FP)
        recall <- TP / (TP + FN)
        F1  <- 2 * (recall * precision) / (precision + recall)
        c(precision = precision,
          recall = recall,
          F1 = F1)
      }
    )
    summaryRules$precision <- performance[1, ] * 100
    summaryRules$recall <- performance[2, ] * 100
    summaryRules$F1 <- performance[3, ] * 100
  }
  return(summaryRules)
}

data <- summarizeRules(subgroup.data = simulated.twoarm, rules = allRules)
data$log_p_value <- -log10(data$p_value)
  
data <- data[data$outcome == "Y_binary",]
ggplot(
  data = data,
  aes(
    x = precision,
    y = recall,
    shape = outcome,
    color = method
  )
) +
geom_point() +
ggtitle("Correctness of rules") +
geom_label(
  label = data$method,
  nudge_x = 0.25,
  nudge_y = 0.2,
  size = 5
)

ggplot(
  data = data,
  aes(
    x = interaction_coeff,
    y = log_p_value,
    shape = outcome,
    color = method,
     size = 10
  )
 
) +
geom_point() +
ggtitle("Volcano plot of rules") +
geom_text(
  label = data$method,
  size = 3
) + 
geom_hline(yintercept = P_THRESHOLD)

data

```



